Title

### 2015-07-26 main.tex
bonnie's edits and reorg

Certainly, computers are getting faster and more economical; the amount of 
processing available per dollar of compute hardware is more or less doubling 
every year or two; a similar claim can be made about storage capacity 
(Figure~\ref{fig:growth}).
In 2002, when the first human genome was sequenced, the growth in computing 
power was still matching the growth rate of genomic data.
However, the sequencing technology used for the Human Genome Project--Sanger
sequencing--was supplanted around 2004, with the advent of what is now known as 
next-generation sequencing.
The material costs to sequence a genome have plummeted in the past
decade, to the point where a whole human genome can be sequenced for less than
one thousand U.S. Dollars.
As a result, by some estimates, the amount of genomic data available to 
researchers is increasing by a factor of 10 every year.

\begin{figure}[htb!]
\centering
\subfloat[Growth of genomic sequence data as compared with the combined power of the top-500 supercomputer list; y-axis is logarithmic]{
\includegraphics[width=3.0in]{assets/moore.png}
\label{fig:moore}
}\\
\subfloat[Growth of genomic sequence data as compared with hard drive capacity; y-axis is logarithmic]{
\includegraphics[width=3.0in]{assets/kryder.png}
\label{fig:kryder}
}
\caption[Sequencing vs. Storage]{Moore's and Kryder's laws contrasted with genomic sequence data}
\label{fig:growth}
\end{figure}

@@

Computer scientists routinely exploit peculiarities in the structure of data in
order to reduce time or space complexity.
In computational biology, this approach has served researchers well.
Now-classical approaches such as principal component analysis (PCA) reduce the 
dimensionality of data in order to simplify analysis and uncover salient 
features.
This review focuses on algorithmic advances for dealing with the growth in biological data; algorithms for novel biological insights are not its focus.

\section{Biological sequence data and next-generation sequencing}

Many, but not all, of the problems in computational biology deal with sequence
data, either nucleotide sequences (nominally a four-letter alphabet) or protein
sequences (nominally a twenty-letter alphabet).
Certainly, as sequencing is generating the greatest volume of biological data,
a large portion of bioinformatics algorithms deal primarily with string data.
For the problem of exact search, approaches for transforming and indexing 
string data provide benefits.
The Burrows-Wheeler transform (BWT)~\cite{burrows1994block} provides efficient string 
compression through a reversible transformation, while the FM-index data 
structure~\cite{ferragina2000opportunistic} is a compressed substring index, based
on the BWT, which provides efficient storage as well as fast search.
Modern next-generation sequencing (NGS) technologies read short fragments of
DNA, known as \emph{reads}~\ref{fig:shotgun}, which must be mapped onto a reference sequence~\ref{fig:ngs-pipeline}
BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
Bowtie2~\cite{langmead2012fast} aligner further relies on the FM-index for efficient 
mapping of NGS reads.

@@

As biological sequence data is string data, it is frequently useful to index or
count instances of $k$-mers of a fixed length, $k$.
Search tools such as BLAST~\cite{altschul1990basic} rely on $k$-mer \emph{seeds} in a 
seed-and-extend approach to sequence search and alignment.
For counting $k$-mer frequencies, Jellyfish~\cite{marccais2011fast} relies on an efficient
hash-table encoding as well as Intel's ``compare-and-swap'' CPU instructions for
improved run-time performance.

Seeking to take advantage of sequence redundancy, \cite{loh2012compressive}
introduced \emph{compressive genomics}, an approach that relies on compressing
data in such a way that the desired computation (such as BLAST search) can be
performed in the compressed representation.
This approach relies on a two-stage search, referred to as \emph{coarse} and \emph{fine} search.
Coarse search is performed only on the coarse, or representative, subsequences
that represent unique data.
Any coarse sequence that is within some threshold of the query is then expanded 
into all the similar sequences it represents; the fine search is then over this
(typically small) subset of the original database.
This approach provides run-time improvements to BLAST search on 
nucleotide~\cite{loh2012compressive} and protein~\cite{daniels2013compressive}
data; moreover, these run-time improvements increase as databases grow.
Also taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
similar reads (those that share a common 
substring) together into ``buckets'', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This allows a general-purpose compressor to achieve tight compression.
SCALCE~\cite{hach2012scalce} also relies on a ``boosting'' scheme, reordering
reads in such a way that a general-purpose compressor achieves improved
compression ratios.

The idea of taking advantage of redundancy in sequence data is particularly
important in the face of new sources of data.
Driven by the plummeting costs of next-generation sequencing, the 1000 Genomes Project~\cite{10002012integrated} is pursuing a broad catalog of human 
variation; instead of producing a single reference genome for a species, many
complete genomes are catalogued.
Likewise, WormBase~\cite{harris2014wormbase} and 
FlyBase~\cite{tweedie2009flybase} are cataloguing many different species and
strains of the \emph{Caenorhabditis} worm and \emph{Drosophila} fruit fly, 
respectively.
On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{gillespie2002isolation,forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease~\cite{gevers2014treatment,manichanh2006reduced}, and 
obesity~\cite{greenblum2012metagenomic}.

One approach to more quickly searching these data sets is to come up with an
indexing technique that leverages domain knowledge about the structure of the
data.
In the realm of NGS read mapping, the Genome Multitool (GEM)
mapper~\cite{marco2012gem} makes use of a Ferragina-Manzini 
index~\cite{ferragina2000opportunistic} coupled with dynamic programming in
a compressed representation, in order to prune the search space when mapping
reads to a reference genome.
Masai~\cite{siragusa2013fast} uses an ``approximate seed'' approach to indexing
the space of possible matches, likewise pruning the search space.

Another approach is to make use of a completely different paradigm for the 
structure of biological data. 
By combining redundancy with a low fractal 
dimension~\cite{yu2015entropy}, it may be possible to develop algorithms that 
scale even as the data continue to grow.
The CORA read mapper~\cite{yorukoglu2015compressive} applies a $k$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
Short seed-clustering schemes, such as those used in Masai and 
MrsFAST~\cite{hach2010mrsfast} conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its k-mer based read compression model, CORA is able to accelerate and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.

In the realm of metagenomic search, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
Recent advances in metagenomic search tools have relied on three advances over
BLASTX: indexing, alphabet reduction, and compression.
RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
collision-free hash table.
The alphabet reduction, as it is reversible, can be thought of as a form of
lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
alphabet, with offsets stored to recover the original sequence in the full
alphabet.
The hash-table provides an efficient index of the database to be searched.
DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
``shaped seeds'' instead of simple $k$-mer seeds to index the database.

The recently-released MICA~\cite{yu2015entropy} demonstrates that the 
compressive-acceleration approach of 
caBLAST~\cite{loh2012compressive} and caBLASTP~\cite{daniels2013compressive} is 
largely orthogonal to alphabet-reduction and indexing approaches.
MICA applies the compressive-acceleration framework to DIAMOND, using DIAMOND
for its ``coarse search'' phase and a user's choice of DIAMOND or BLASTX for its
``fine search'' phase; it demonstrates comparable run-time gains over DIAMOND as
caBLASTP does over BLASTP.

\section{Non-sequence data}

@@

In the realm of protein structure prediction, it has been 
observed~\cite{dunbrack2002rotamer} that rather than exploring the continuous 
range of
possible dihedral angles, protein structures inhabit a smaller subspace, defined
by a narrower range of angles, referred to as 
rotamers~\cite{dunbrack1993backbone}.
The TreePack algorithm~\cite{xu2006fast} takes advantage of this observation 
for the
purpose of sidechain packing (finding the lowest-energy conformation for the
sidechain of each amino acid), using tree width to reduce the search space and
simplify dependencies.

@@

When the structure of a protein is known but its biological function or 
evolutionary relationships are not, researchers may search for structurally
similar proteins that are better studied~\cite{gibrat1996surprising}.
Classical tools for this involve performing pairwise structural alignments to
look for geometric similarity; DALI~\cite{holm1995dali} is still widely used,
along with other aligners such as FATCAT~\cite{ye2004fatcat} and 
Matt~\cite{menke2008matt}.
FragBag~\cite{budowski2010fragbag} accelerates protein structure search by
approximating structural alignments, which can be computationally demanding,
by instead comparing the ``bag-of-words'' from each structure.
Analogous to a term-frequency vector in information retrieval, this bag-of-words
indicates the abundance of particular, short structural motifs within a protein.
Again applying a compressive acceleration approach, 
esFragBag~\cite{yu2015entropy} clusters proteins based on the cosine distance or
Euclidean distance of their bag-of-words vectors, further accelerating FragBag's
running time by an average of 10x.

@@

% Bibliography
% \bibliographystyle{ACM-Reference-Format-Journals}
\bibliographystyle{abbrv}
\bibliography{main}

>>>>

Certainly, computers are getting faster and more economical; the amount of 
processing available per dollar of compute hardware is more or less doubling 
every year or two; a similar claim can be made about storage capacity 
(Figure~\ref{fig:growth}).
In 2002, when the first human genome was sequenced, the growth in computing 
power was still matching the growth rate of genomic data.
However, the sequencing technology used for the Human Genome Project--Sanger
sequencing--was supplanted around 2004, with the advent of what is now known as 
next-generation sequencing.
The material costs to sequence a genome have plummeted in the past
decade, to the point where a whole human genome can be sequenced for less than
one thousand U.S. Dollars.
As a result, the amount of genomic data available to 
researchers is increasing by a factor of 10 every year.

\begin{figure}[htb!]
\centering
\subfloat[Growth of genomic sequence data (orange) as compared with the combined power of the top-500 supercomputer list (blue); y-axis is logarithmic]{
\includegraphics[width=3.0in]{assets/moore.png}
\label{fig:moore}
}\\
\subfloat[Growth of genomic sequence data (orange) as compared with hard drive capacity (blue); y-axis is logarithmic]{
\includegraphics[width=3.0in]{assets/kryder.png}
\label{fig:kryder}
}
\caption[Sequencing vs. Storage]{Moore's (a) and Kryder's (b) laws contrasted with genomic sequence data}
\label{fig:growth}
\end{figure}

@@

Computer scientists routinely exploit peculiarities in the structure of data in
order to reduce time or space complexity.
In computational biology, this approach has served researchers well.
Now-classical approaches such as principal component analysis (PCA) reduce the 
dimensionality of data in order to simplify analysis and uncover salient 
features~\cite{berger2013computational}.
As another example, clever indexing techniques such as the Burroughs-Wheeler 
Transform (BWT) also take advantage of aspects of sequence 
data~\cite{berger2013computational}.
This review focuses on algorithmic advances for dealing with the growth in 
biological data; algorithms for gaining novel biological insights are not its 
focus.

\section{Types and sources of biological data}

In the central dogma of molecular biology, DNA is transcribed into RNA, which
is translated by the ribosome into a polypeptide chain, a sequence of amino 
acids, known as a protein.
This protein folds into a complex, low-energy structure, and functions as
a cellular machine itself; the DNA sequence determines the amino acid sequence,
which in turn determines the folded structure of the protein.
This structure ultimately determines the protein's function within the cell.
Certain kinds of RNA also function, sometimes with proteins, as cellular 
machines.
Thus, researchers are interested in DNA, RNA, and protein sequences, as well
as the structures that RNA and proteins form.
Researchers also study the resulting functions of RNA and proteins, in what
abundance they occur (this is known as gene expression), and how they interact
with each other (interaction networks).

Most but not all biological data take the form of \emph{sequence} data, either 
nucleotide sequences (nominally a four-letter alphabet representing the four 
DNA or RNA bases) or protein sequences (nominally a twenty-letter alphabet 
representing the twenty standard amino acids).
Sequence data are obtained in several ways, including mass spectrometry (which
can determine protein sequence) and RNA-seq (which can determine RNA sequence, 
and by inference the protein sequence into which it might translate).
However, the greatest volume of sequence data derive from next-generation 
sequencing (commonly abbreviated NGS) technologies, and comprise DNA sequences.
At the dawn of the genomic era, Sanger sequencing was the most widely used 
method.
Slow and labor intensive but well-tested, Sanger sequencing is still used to 
validate NGS results.
More recently, Sanger sequencing has largely been supplanted by NGS approaches,
beginning with Illumina's ``sequencing by synthesis,'' which allows vastly 
greater throughput due to massive parallelism, low cost, and simple preparation
of samples.
Importantly to this Review, Illumina sequencing, as well as other approaches
such as SOLiD, Ion Torrent, and 454 pyrosequencing, does not read a single DNA
molecule end-to-end as one could a file from a hard drive.
Instead, DNA molecules are chopped into many small fragments, or 
\emph{reads} (Figure ~\ref{fig:shotgun}), 
which are read concurrently, and which must be reassembled in the correct order 
if an entire genome is to be read.
Reads most typically range from 50 to 200 bases long, though longer reads are
available with some technologies.
This so-called ``shotgun sequencing'' is analogous to the problem of 
reconstructing a large file from small, 
randomly-ordered reads, or a shredded paper document.
However, these NGS reads overlap, and they are many in number, so 
reconstructing an entire genome is feasible.
No sequencing technology is completely infallible, so sequencing machines also
provide a \emph{quality score} (or measure of the confidence in the DNA base
called) associated with each position.
Thus, an NGS read is a string of DNA letters, coupled with a string of ASCII
characters that encode the quality of the base call.
A sequencing run will produce many of these reads, typically stored in a FASTQ
file.

Protein structure, primarily determined through X-ray crystallography or
nuclear magnetic resonance (NMR), describes the coordinates of each atom, in 
Angstroms, of a protein in three-dimensional space.
This information is important to understanding how a protein functions, because
proteins interact physically with other molecules.
Protein structure data are commonly stored in the Protein Data Bank (PDB); 
each PDB entry has an associated resolution indicating the accuracy of the
structural description.
Thus a PDB entry at its most basic level contains a list of atoms, along with
which amino acid they belong to and their spatial coordinates as floating-point
numbers.

Chemical compounds represent another source of structural data, more general
than proteins.
Determined by X-ray crystallography, NMR, electron microscopy and other 
techniques, compounds---many of which are useful pharmaceuticals---can be
represented as labeled graphs.
In such a graph, the vertices are atoms (labeled with chemical element) and
the edges are bonds between atoms (labeled with bond type, such as single or
double).
Chemical compounds are stored in databases such as PubChem, and can be described
as adjacency matrices.

\section{Challenges in biological data}

Given DNA or RNA reads from NGS technologies, the first task is to 
\emph{assemble} those fragments of sequence (which have associated quality 
scores) into contiguous sequences.
The task of \emph{de novo} assembly is beyond the scope of this Review, but is
possible because reads overlap and because the sequence is covered by many such
overlapping reads~\cite{zhang2011practical}; the \emph{de Bruijn graph} is a 
commonly-used data structure.
Often, however, a \emph{reference genome} (or in the case of RNA, 
\emph{transcriptome}) is available for the organism being sequenced; the
establishment of a human reference genome was indeed the purpose of the Human
Genome Project.
When a reference sequence is available, NGS reads must be \emph{mapped} onto
this reference (Figure ~\ref{fig:ngs-pipeline}).
This allows the differences between the newly-sequenced genome and the reference
to be analyzed; these differences may include single-nucleotide polymorphisms
(SNPs, which are the genetic analogue to bit-flips) (Figure~\ref{fig:snps}), 
insertions 
or deletions, or larger-scale changes in the genome.
While reference-based read mapping is a fundamentally simpler problem than
\emph{de novo} assembly, it is still computationally complex, as gigabytes or
terabytes of reads must each be mapped onto the reference genome, which can 
range from millions (for bacteria) to billions (for mammals).

Standard techniques for reference-based read mapping rely on algorithmic 
approaches such as the Burrows-Wheeler transform (BWT)~\cite{burrows1994block}, 
which provides efficient string 
compression through a reversible transformation, while the FM-index data 
structure~\cite{ferragina2000opportunistic} is a compressed substring index, 
based on the BWT, which provides efficient storage as well as fast search.
BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
Bowtie2~\cite{langmead2012fast} aligner further relies on the FM-index for 
efficient mapping of NGS reads.
The Genome Multitool (GEM) mapper~\cite{marco2012gem} also uses an FM-index 
coupled with dynamic programming in a compressed representation, in order to 
prune the search space when mapping reads to a reference genome.
Masai~\cite{siragusa2013fast} uses an ``approximate seed'' approach to indexing
the space of possible matches, likewise pruning the search space.
Even with these approaches, read mapping remains a significant bottleneck in
genomic research~\cite{hach2010mrsfast}.

@@

Given a sequenced genome, the next natural questions are what genes (genomic
regions that code for proteins) are present, what structure each resulting
protein takes, and what biological function it performs.
Identifying likely genes is a well-studied problem~\cite{burge1998finding} 
beyond the scope of this Review.
However, determining evolutionary relationships, structure, and function are
at the heart of current research in computational biology.
Since some organisms (known as \emph{model organisms}) are better-studied than
others, and evolution is known to conserve sequence, structure, and function,
a powerful approach to determine these attributes is to search for similar
sequences about which more is known.
This so-called \emph{homology search} entails searching for \emph{approximate}
matches in databases of known gene or protein sequences.

Basic Local Alignment Search Tool (BLAST)~\cite{altschul1990basic} has been
the standard too for performing homology search on databases of nucleotide and
protein sequences.
BLAST takes a ``seed-and-extend'' approach; it looks for small, $k$-mer matches
that might lead to longer matches, and greedily extends them, ultimately 
producing a \emph{sequence alignment} between a query and each potential 
database hit.
However, BLAST's running time is linear in the size of the database being
searched, which is problematic as sequence databases continue to grow at a
faster rate than Moore's law.

Driven by the plummeting costs of next-generation sequencing, the 1000 Genomes Project~\cite{10002012integrated} is pursuing a broad catalog of human 
variation; instead of producing a single reference genome for a species, many
complete genomes are catalogued.
Likewise, WormBase~\cite{harris2014wormbase} and 
FlyBase~\cite{tweedie2009flybase} are cataloguing many different species and
strains of the \emph{Caenorhabditis} worm and \emph{Drosophila} fruit fly, 
respectively.
On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{gillespie2002isolation,forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease~\cite{gevers2014treatment,manichanh2006reduced}, and 
obesity~\cite{greenblum2012metagenomic}.

Metagenomics fundamentally asks what organisms are present, and (in the case
of a microbiome such as the gut) what metabolic functions it can accomplish as
a whole.
One approach to this is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second approach is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
Recent advances in metagenomic search tools have relied on two advances over
BLASTX: indexing and alphabet reduction.
RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
collision-free hash table.
The alphabet reduction, as it is reversible, can be thought of as a form of
lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
alphabet, with offsets stored to recover the original sequence in the full
alphabet.
The hash-table provides an efficient index of the database to be searched.
DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
``shaped seeds'' instead of simple $k$-mer seeds to index the database.
DIAMOND demonstrates search performance three to four orders of magnitude faster
than BLASTX, but it is still linear in the size of the database being searched.

Over the course of evolutionary time, structure is known to be more highly 
conserved than sequence~\cite{illergaard2009structure}, which is to say that it 
does not change as rapidly.
When the structure of a protein is known but its biological function or 
evolutionary relationships are not, researchers may search for structurally
similar proteins that are better studied~\cite{gibrat1996surprising}.
Classical tools for this involve performing pairwise structural alignments to
look for geometric similarity; DALI~\cite{holm1995dali} is still widely used,
along with other aligners such as FATCAT~\cite{ye2004fatcat} and 
Matt~\cite{menke2008matt}.
FragBag~\cite{budowski2010fragbag} accelerates protein structure search by
approximating structural alignments, which can be computationally demanding,
by instead comparing the ``bag-of-words'' from each structure.
Analogous to a term-frequency vector in information retrieval, this bag-of-words
indicates the abundance of particular, short structural motifs within a protein.

@@

\section{Compressive omics}

Another approach is to make use of a completely different paradigm for the 
structure of biological data.
Much biological data is highly redundant; if a computation is performed on one
human genome, and a researcher wishes to perform the same computation on another
human genome, most of the work has already been done.

Seeking to take advantage of the redundancy inherent in genomic sequence data, 
\cite{loh2012compressive}
introduced \emph{compressive genomics}, an approach that relies on compressing
data in such a way that the desired computation (such as BLAST search) can be
performed in the compressed representation.
This approach relies on a two-stage search, referred to as \emph{coarse} and \emph{fine} search.
Coarse search is performed only on the coarse, or representative, subsequences
that represent unique data.
Any coarse sequence that is within some threshold of the query is then expanded 
into all the similar sequences it represents; the fine search is then over this
(typically small) subset of the original database.
This approach provides run-time improvements to BLAST search on 
nucleotide~\cite{loh2012compressive} and protein~\cite{daniels2013compressive}
data; moreover, these run-time improvements increase as databases grow.

The CORA read mapper~\cite{yorukoglu2015compressive} applies an $l$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
CORA, like caBLAST and caBLASTP, accelerates existing tools (in this case, read
mappers including BWA, and Bowtie2) by allowing them to operate in a compressed
space, and relying on a coarse and a fine phase.
Short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
MrsFAST~\cite{hach2010mrsfast} conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its k-mer based read compression model, CORA is able to accelerate and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.

NEED TO DISCUSS $l$-mers here.

In the area of metagenomic search,
the recently-released MICA~\cite{yu2015entropy} demonstrates that the 
compressive-acceleration approach of 
caBLAST~\cite{loh2012compressive} and caBLASTP~\cite{daniels2013compressive} is 
largely orthogonal to alphabet-reduction and indexing approaches.
MICA applies the compressive-acceleration framework to 
DIAMOND~\cite{buchfink2014fast}, using DIAMOND
for its ``coarse search'' phase and a user's choice of DIAMOND or BLASTX for its
``fine search'' phase; it demonstrates comparable run-time gains over DIAMOND as
caBLASTP does over BLASTP.

@@

Again applying a compressive acceleration approach, 
esFragBag~\cite{yu2015entropy} clusters proteins based on the cosine distance or
Euclidean distance of their bag-of-words vectors, further accelerating FragBag's
running time by an average of 10x.

\section{Structure of biological data}

Cluster-based search is well-studied~\cite{jardine1971use}, but conventional
wisdom holds that it provides a constant factor speed-up over exhaustive search.
However, \cite{daniels2013compressive} observed that caBLASTP demonstrated
increased gains over BLASTP on newer, larger protein databases, suggesting that
compressive acceleration takes advantage of ever-increasing redundancy in the
data.
Beyond redundancy, however, another attribute of large biological data sets
stands out.
Any high-dimensional data set, such as amino acid sequences of a bounded 
length, can be described as inhabiting a vector space.
For example, given an alphabet of 20 amino acids and a length bounded at 
500\footnote{While the largest-known protein, titin, is over 33,000 amino acids 
in length, proteins are typically described as comprising \emph{domains} that 
evolve and fold independently; domains rarely exceed 500 amino acids.}, one 
could
enumerate $20^{500} \approx 3 \times 10^{650}$ distinct protein sequences, as 
of July 
2015, the NCBI's exhaustive ``NR'' database lists $6.9 \times 10^{7}$ sequences.
This is not simply due to undersampling; thanks to evolution, only those protein
sequences which exhibit useful biological function should survive, and most
random sequences of amino acids would not be expected to form stable structures.
Thus, not only do biological data exhibit redundancy, they also tend not to 
inhabit the entire feasible space.
It seems that physical laws have constrained the data to a particular subspace
of the enumerable Cartesian space.

One key insight as to why the ``compressive omics'' approach allows computation
that scales sublinearly in the size of the data is that such data sets exhibit
low \emph{metric entropy}~\cite{yu2015entropy}.
That is, for a given cluster radius $r_c$ and a database $D$, the number 
$N_{r_c} (D)$ of clusters needed to cover $D$ is relatively small compared to
$|D|$, the number of entries in the database.
In contrast, if the points were uniformly distributed about the Cartesian space,
$N_{r_c} (D)$ would be larger.

A second key insight is that the biological data sets considered 
by~\cite{yu2015entropy} exhibit low \emph{fractal dimension}.
That is, for some range of radii $r_1$ and $r_2$,

discrete?

By combining redundancy with a low fractal 
dimension~\cite{yu2015entropy}, it may be possible to develop algorithms that 
scale even as the data continue to grow.

wherever physical or empirical laws have constrained data to 
a subspace of low entropy and fractal dimension.

Specifically, the ratio $\frac{|D|}{k}$ provides an estimate of the acceleration factor
for just the coarse search component compared to a full linear search.
Local fractal dimension around a data point can be computed by determining the
number of other data points within two radii $r_1$ and $r_2$ of that point;
given those point counts ($n_1$ and $n_2$, respectively), fractal dimension $d$
is simply $d=\frac{\log (n_2 / n_1)}{ \log (r_2 / r_1)}$.
Sampling this property over a dataset can provide a global average fractal 
dimension.
When we search a larger radius around a query, the number of points we encounter grows exponentially with the fractal dimension;
low fractal dimension implies that this growth will not obviate the gains provided by an entropy-scaling data structure.

An entropy-scaling search tree is also related to a metric ball tree~\cite{uhlmann1991satisfying}, although with different time complexity.
Querying a metric ball tree requires $O(\log n)$ time, assuming the relatively uniform distribution of data points in a metric space.
This is different from the nonuniform distribution under which entropy-scaling
search behaves well.

\section{Storage compression}

Also taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
similar reads (those that share a common 
substring) together into ``buckets'', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This allows a general-purpose compressor to achieve tight compression.
SCALCE~\cite{hach2012scalce} also relies on a ``boosting'' scheme, reordering
reads in such a way that a general-purpose compressor achieves improved
compression ratios.

% In the realm of protein structure prediction, it has been
% observed~\cite{dunbrack2002rotamer} that rather than exploring the continuous
% range of
% possible dihedral angles, protein structures inhabit a smaller subspace, defined
% by a narrower range of angles, referred to as
% rotamers~\cite{dunbrack1993backbone}.
% The TreePack algorithm~\cite{xu2006fast} takes advantage of this observation
% for the
% purpose of sidechain packing (finding the lowest-energy conformation for the
% sidechain of each amino acid), using tree width to reduce the search space and
% simplify dependencies.

@@

% Bibliography
\bibliographystyle{ACM-Reference-Format-Journals}
% \bibliographystyle{abbrv}
\bibliography{main}
